---
title: "Bayesian Changepoint Detection on Price Histograms"
author: "Pavel Logačev"
date: "r Sys.Date()"
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
    df-print: paged
engine: knitr
editor: visual
---

```{r message=FALSE, warning=FALSE}
library(Rcpp)
source("./source/data_generation.r")
source("./source/greedy_cp_selection.r")
Rcpp::sourceCpp("./source/greedy_cp_selection.cpp")
```

## Summary

This notebook demonstrates a Bayesian changepoint detection algorithm for histogram-valued time series. It is based on a greedy search  
tailored for transactional price data with varying pricing regimes.

-   Each day’s price distribution is modeled via discretized histograms.
-   The changepoint configuration is selected via greedy MAP estimation.
-   In the future, parameters may be obtained by means of sampling instead of greedy search. (For instance, using RcppSMC or a custom sampler, e.g., MH).

## Modeling Approach

Each day is represented by a histogram over discrete price points. To this end, the time series is transformed into a histogram-valued sequence, 
represented as a matrix of counts.
We assume that pricing follows certain *pricing regimes*, each associated with a different price distribution. Changes in pricing regime occur when 
the pricing of a product undergoes a meaningful change — such as list price changes, promotions, or changes in the availability of discounts.
Changepoints are defined as time indices where the underlying price distribution regime changes, leading to a change in relative frequency of price points.


### Data Format

In the synthetic data set used here, we assume multiple pricing regimes, each characterized by a different list price, a different discount price, or 
a different discount probability. We generate a synthetic data set with 800 days, with 9 changepoints, with the following properties:

```{r}
set.seed(123)

segments <- data.frame(
             start_day = c(1,  90, 250, 300, 400, 500, 550, 600, 700, 750),
            list_price = c(5,    6,   6,   6,   6, 5.5, 5.5, 5.5, 7, 7),
        discount_price = c(4,    4,   5,   5,   5, 4.5, 4.5, 4.5, 2, 2),
  discount_probability = c(.25, .1,  .1, .25,  .5, .01, .03, .15,.3, .25)
)

n_days <- 800
lambda_qty <- 20
df <- generate_transaction_prices(segments, n_days, lambda_qty, seed = 123)
hist <- compute_price_histogram(df)
actual_changepoints <- segments$start_day[-1]-0.5

knitr::kable(segments, caption = "Price Regimes Overview")
```

The plot below shows the synthetic data set, with the actual changepoints marked with dashed vertical lines.

```{r fig.height=3, fig.width=10}
p_price <- hist$df %>% rename(quantity = qty) %>%  filter(quantity > 0) %>%
    ggplot(aes(day, price)) + geom_point(aes(size=quantity, alpha = quantity)) + 
    geom_vline( data = data.frame(x = actual_changepoints, changepoint = "actual"),
                aes(xintercept = x, color = changepoint), linetype = "dashed") + theme_bw() +
    theme(legend.position = "top") 

print(p_price)
```

### Likelihood and Posterior

We define the likelihood of a segmentation as the sum of within-segment log-likelihoods, each modeled as a multinomial distribution over price bins.
To regularize the solution and avoid overfitting, we place a Bernoulli prior on each possible changepoint with success probability λ.
The log-posterior of a changepoint configuration is:

$$
 \text{log}~p(z∣data) \propto \text{log}~p(data∣z) + \text{log}~p(z)
$$


The core of the model is the log-likelihood of a price histogram segment. 
For a time window $[t_1, t_2]$, we compute the (regularized) empirical histogram:

```{=html}
<pre><code class="language-cpp">
double log1m_exp(double x)
{
if (x >= 0.0) stop("log1m_exp is undefined for x >= 0");
...
}
</code></pre>
```

Each histogram is treated as a multinomial count vector. We compute the segment's likelihood using the empirical distribution’s entropy:

$$
\log p(\mathbf{n}) = \sum_{i=1}^{K} n_i \log\left( \frac{n_i + \epsilon}{\sum_j (n_j + \epsilon)} \right)
$$

This avoids instability from zero counts via additive smoothing (`epsilon`).



## Estimation

To estimate the changepoint configuration c, we use a greedy forward search:

1.  Start with no changepoints.
2.  Iteratively add the changepoint that most increases the penalized log-likelihood.
3.  Stop when no further improvement is possible.

This process is repeated for various values of $\lambda$ (transformed from a logit_lambda scale), and the $\lambda$ value that maximizes the resulting log-posterior is selected using one-dimensional optimization (`optimize()` in R).

## Algorithm and Implementation

This project implements a greedy changepoint detection algorithm for time series of price histograms. The codebase consists of three components:

-   A data generator that simulates daily transaction-level price data under piecewise constant pricing regimes (`generate_transaction_prices()`).
-   A transformation step that maps transactional data into a histogram matrix (`compute_price_histogram()`).
-   A C++ backend that performs fast log-likelihood evaluation and greedy changepoint selection via Rcpp.

## Segment Likelihood: Histogram Log-Likelihood

## Data Generation


## Changepoint Detection

```{r }
opt_res <- locate_optimal_changepoints( hist$histogram, max_lambda = 0.3 )
detected_changepoints <- which(opt_res$changepoints) - 0.5
detected_changepoints
```

```{r fig.height=3, fig.width=10}
p_price +
      geom_vline( data = data.frame(x = detected_changepoints, changepoint = "detected"),
                  aes(xintercept = x, color = changepoint), linetype = "dashed") +
      scale_color_manual(name = "changepoint", values = c("detected" = "blue", "actual" = "red"),
                        guide = guide_legend(override.aes = list( linetype = "solid", size = 1))
                        )
```

## Repository

All source code is available here:\
👉 <https://github.com/plogacev/case_studies/tree/main/pricedist_changepoints>